\documentclass[notes,color]{sepslide0}
\usepackage{graphicx}
\usepackage[overheads]{mysepslides}
\usepackage{tech,graphicx,url,tikz,scalalistings}

\title{Semaphores} 
\author{Gavin Lowe}

\everymath{\color{violet}}
\def\scalacolour{\color{violet}}
\def\codecolour{\scalashape}

\begin{document}

\begin{slide}
  
  \Title

Reading: Andrews Chapter 4.
\end{slide}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{semaphores1} % introduction, use for mutex
\input{semaphores2} % signalling, passing the baton, queue, barrier,
                    % queue-lock
% \input{semaphores-syncchan}
\input{semaphores-readers-writers}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%%%%%%%%%

\begin{slide}
\heading{Counting semaphores}

The semaphores we have seen so far have been binary: they have just two
states, up and down.

A counting semaphore has a non-negative integer as its state.
%
\begin{itemize}
\item
The \SCALA{down} operation waits until the value is strictly positive, and
decrements it;

\item
The \SCALA{up} operation increments the value, waking a process if necessary.
\end{itemize}
%
The value of the semaphore can be thought of as the number of \emph{permits}
available for |down| operations. 
\end{slide}

%%%%%

\begin{slide}
\heading{Counting semaphores}

\begin{scala}
class CountingSemaphore(private var permits: Int = 0){

  def down = synchronized{
    while(permits == 0) wait()
    permits -= 1
  }

  def up = synchronized{
    permits += 1
    notify()
  }
}
\end{scala}
\end{slide}

%%%%%

\begin{slide}
\heading{A partial queue}

We can implement a queue, using a counting semaphore to block attempts to
remove elements until the buffer is non-empty.

\begin{scala}
/** A queue implemented using a counting semaphore. */
class CountingSemaphorePartialQueue[T] extends PartialQueue[T]{
  /** The queue itself. */
  private val queue = new scala.collection.mutable.Queue[T]

  /** Semaphore for mutual exclusion on queue. */
  private val mutex = new MutexSemaphore

  /** Semaphore for dequeueing.  The state of the semaphore equals
    * queue.length. */
  private val size = new CountingSemaphore(0)
  ...
}
\end{scala}
\end{slide}

%%%%%


\begin{slide}
\heading{A partial queue}

\begin{scala}
  def enqueue(v: T) = {
    mutex.down
    queue.enqueue(v)
    size.up
    mutex.up
  }

  def dequeue: T = {
    size.down
    mutex.down
    val result = queue.dequeue
    mutex.up
    result
  }
\end{scala}
\end{slide}

%%%%%



\begin{slide}
\heading{A partial queue}

Note that the order in which |dequeue| obtains the semaphores is
important.  If it were to obtain |mutex| first, and |queue| is empty,
then this would block |enqueue|, so the system would be deadlocked. 

\emph{Exercise:} adapt this example to produce a bounded buffer that cannot
hold more than $n$ pieces of data.
\end{slide}


% \begin{scala}
% class Buff[T]{
%   private val queue = new scala.collection.mutable.Queue[T];
%   private val mutex = new Semaphore; 
%   private val size = new CountingSemaphore(0);
%   ...
% }
% \end{scala}
% %
% \SCALA{queue} holds the data.  

% \SCALA{mutex} is used to ensure mutual exclusion of operations upon
% \SCALA{queue}.  

% \SCALA{size} is a counting semaphore which stores the number of elements
% currently in \SCALA{queue}; it is used to block attempts to remove elements
% when its value is~$0$.
% \end{slide}

% %%%%%

% \begin{slide}
% \heading{An unbounded buffer}

% \begin{scala}
% class Buff[T]{
%   ...

%   def add(v:T) = {
%     mutex.down;
%     queue.enqueue(v);
%     size.up; mutex.up;
%   }

%   def remove:T = {
%     size.down; mutex down;
%     val result = queue.dequeue;
%     mutex.up; return result;
%   }
% }
% \end{scala}
% \end{slide}

%%%%%

% \begin{slide}
% \heading{A bounded buffer}

% \emph{Exercise:} adapt this example to produce a bounded buffer, that cannot
% hold more than $n$ pieces of data.
% \end{slide}

%%%%%

\begin{slide}
\heading{Summary}

\begin{itemize}
\item
Semantics of semaphores;

\item
Implementing a semaphore using a monitor;

\item
Using a semaphore to enforce mutual exclusion;

\item
Dangers of deadlock;

\item
Using a semaphore for signalling;

\item
Examples;

\item
Passing the baton;

\item
Fairness;

\item
Counting semaphores.
\end{itemize}
\end{slide}

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
