\begin{question}
Consider the following synchronisation problem.  A savings account is shared
by $N$ people, each represented by a process.  Each process can make
deposits and/or withdrawals.  The account records the current balance; this
may never go negative.  As a result, a withdrawal may be blocked until
sufficient funds become available.  

The aim of this question is to implement the account using the three different
types of concurrency primitive: message passing, monitors and semaphores.  In
each part of the question you should use only the concurrency primitive
mentioned. 

Throughout this question, the system does not need to be fair as to which
pending deposit is unblocked once sufficient funds become available.

%%%%%%

\begin{enumerate}
\item
Implement the account using a server process with message passing.  Requests
from clients should be made using messages of type \SCALA{Request}:
\begin{scala}
  type ClientId = Int;
  type Amount = Int;
  abstract class Request;
  case class Deposit(n:Amount, c:ClientId) 
         extends Request;
  case class Withdrawal(n:Amount, c:ClientId) 
         extends Request;
\end{scala}
%
Assume that the clients share a channel \SCALA{req} on which requests are
made.  In response to requests, the server should send a response giving the
current balance; assume each client \SCALA{c} has its own response channel
\SCALA{resp(c)}.  Hence the signature of the server process is:
%
\begin{scala}
  def Server(req: ?[Request], resp: Seq[![Int]]) 
  = proc{ ... }
\end{scala}
%
(You do not need to implement the clients.)
                                                                \marks{8}

%%%%%

\item
Now implement the account as a monitor with the following signature:
%
\begin{scala}
object Account{
  def Credit(amount:Amount, c:ClientId) : Int = ...
  def Debit(amount:Int, c:ClientId) : Int = ...
}
\end{scala}
                                                               \marks{6}

%%%%%

\item
Finally, implement the account using semaphores; you should use an object with
the same signature as in the previous part.

{\bf Hint:} you might find it useful to use an array of semaphores, one for
each client process, on which it waits when a withdrawal is blocked.
                                                               \marks{11}
\end{enumerate}
\end{question}

%%%%%%%%%%

\begin{answer}\null
{\bf Part a.}
%\Footnotesize

\begin{scala}
def Server(req: ?[Request], resp: Seq[![Int]]) = proc{
  var balance = 0; // current balance

  val withdrawals = new Array[Int](5); 
    // amounts of queued withdrawals; -1 represents nothing queued
  for(i <- 0 until N) withdrawals(i) = -1;

  repeat(
    req? match {
      case Deposit(n,c) => {
        balance += n;
        resp(c)!balance;
        // Try to unblock withdrawals
        for(i <- 0 until N)
          if(withdrawals(i)>0 && withdrawals(i)<=balance){
            balance -= withdrawals(i); withdrawals(i) = -1; 
            resp(i)!balance;
          }
      }
      case Withdrawal(n,c) => 
        if(n<=balance){ balance -= n; resp(c)!balance; }
        else withdrawals(c) = n;
    }
  )
}
\end{scala}

\noindent
{\bf Part b.}
%
\begin{scala}
object Account{
  private var balance = 0; // current balance

  def Credit(amount:Int, c:ClientId) : Int = synchronized{
    balance += amount;
    notifyAll();
    return balance;
  }

  def Debit(amount:Int, c:ClientId) : Int = synchronized{
    while(balance<amount) wait();
    balance -= amount;
    return balance;
  }
}
\end{scala}

\noindent
{\bf Part c.}

\begin{scala}
object Account{
  private var balance = 0; // current balance

  // Semaphores and amounts for pending withdrawals
  val sems = new Array[Semaphore](N);
  val amounts = new Array[Int](N); 
  for(i <- 0 until N){ 
    sems(i) = new Semaphore; sems(i).down; amounts(i) = -1; 
  }

  val mutex = new Semaphore; // For mutual exclusion

  // Wake up a waiting process, or lift mutex
  private def Signal = {
    var n = 0;
    // Find process to wake up
    while(n<N && (amounts(n)<0 || amounts(n)>balance)) n += 1;
    if(n<N){ // wake up waiting debit
      amounts(n) = -1; sems(n).up;
    }
    else mutex.up;
  }

  def Credit(amount:Int, id:Int) : Int = {
    mutex.down;
    balance += amount; val result = balance;
    Signal;
    return result;
  }

  def Debit(amount:Int, id:Int) : Int = {
    mutex.down;
    var result = 0;
    if(balance<amount){
      amounts(id) = amount;
      mutex.up; sems(id).down; // wait to be woken
      assert(amount <= balance);
      // Can now proceed
      balance -= amount; result = balance;
      Signal;
    }
    else{ // Can proceed immediately
      balance -= amount; result = balance;
      mutex.up; // No need to signal
    }
    return result;
  }
}
\end{scala}
\end{answer}
