\documentclass[fleqn,12pt]{article}
\usepackage{bigpage,tech,url,javalistings,graphicx,csp-cm}
\usepackage{questions}
\def\color#1{}
\def\Programming{\textbf{[Programming]}}

\Small
%\questionsonly
\questionsandanswers
%\answersonly
%\nontuteanswersonly

\title{Concurrent Programming: Exercises}
\author{Gavin Lowe}

\sloppy

\begin{document}
\maketitle

You should complete all questions marked \Programming\ at a computer, and
include suitable testing code and test results with your answers.

Note that not all exam questions are included here.


\section{Introduction}

\input{web-browser} % Why make a web browser use concurrency?  

\input{mvc} % omit

%% \framebox{***} Why should a program with a GUI use concurrency?

%% \framebox{***} Java event model



\input{interleavings1} % Count number of execution seqs - formula

\input{interleavings2} % based on Andrews 2.10; results of particular
                       % interleaving 

\input{interleavings3} % based on Andrews 2.15

\input{queue} %Independent actions of queue -- add, remove, isFull, isEmpty --
              %what can go wrong? Doesn't really work.

\input{accountRace}

\input{CSPVsCSOEvents}

\framebox{Problems with Google calendar}

%%%%%

\section{Channels}

\input{mergestreams} % merge two ordered streams

\input{integrator}

% \input{mergestreams2} % merge two ordered streams, which could be closed

\input{pipeSort} % *** sort list of N numbers using pipe, A7.3

%%%%%

\section{Parallel programming}

\input{nonDividing} % Remove assumption that \# processors divides \# segments.

\input{matrix} % Matrix multiplication

\input{countDups}
% Given two arrays a and b, without repetitions, count the number of distinct
% elements that appear in both a and b. 

% \input{countDups2} % Same but give full code

\framebox{Travelling salesman?}  Andrews 9.7

\input{pi} % *** Estimating pi: pick random numbers x, y in [0,1);  prob$(x^2+y^2)<1 = \pi/4$. 

%%%%%


\section{Alternation}

\input{merge} 

\input{changeMachine}

\input{altVsExtchoice}

\input{altRestriction}

\input{unboundedBuff}

\input{resettableIntegrator}

\framebox{Traffic lights?}


%%%%%

\section{Clients and Servers} % A7.6: Readers \& Writers 

\input{readersWriters}

\input{menWomenServer}

\framebox{Elevator algorithm?}

\framebox{Several-caster} 2008 exam, Q2

\framebox{Multi-casting connections} 2008 exam, Q3

\framebox{Consider} a set of processes where each process is identified by
a unique integer. Each process has a constrained region, which can be entered
only if the sum of all integers associated with processes already executing in
their constrained regions is divisible by three.  Implement this in a
client-server pattern.  Is it possible for a process to be locked out?

\input{savingsAccount} % A7.8 Savings account problem

\input{oneLaneBridge} % ***? A7.11 One-lane bridge

%%%%%

\section{Interacting Peers}

\input{maleFemale} % based on A7.9 under interacting peers.

\input{ringFold} % calculate fold round ring

% \input{smallestCommon} % ***(part) A7.16 smallest common int

\framebox{A7.17 distributed pairing} -- needs a symmetry breaker?

\framebox{Decide conjunction of boolean variables held by nodes} -- with
optimisation -- and then polymorphic generalisation.

\framebox{Drinking Philosophers:} ***
\url{http://www.cs.utexas.edu/~misra/scannedPdf.dir/DrinkingPhil.pdf} -- MSc
exam? 

\input{distTermination} % Distributed termination

%%%%%%%%%%%%%%%%

\section{Parallel computation}

\input{barrierLog} % Implement barrier synchronisation in O(log N) steps

\input{prefixSum} % In the prefix sum example, what can go wrong without the
                  % barrier synchronisation?

\input{prefixSumShared} % Prefix sum with shared variables

\input{JacobiMP} % Convert Jacobi iteration into message-passing program.
% Now done in lectures

\framebox{LaPlace equation}

\input{regionLabelling} % ***  Region labelling problem (A 3.26, 3.27), using
% message passing, termination signalled via messages ***

\framebox{Smoothing (A 9.9)}

\framebox{Game of life on unbounded board}; keep track of coords of live cells.

\framebox{Model motion of snooker balls.}

\framebox{Monte carlo simulation??}

\input{graphSearch} % *** Breadth-first search in a graph

\input{gridMax}
% *** $N^2$ processes, each has an integer; each should end up with the sum of
% all the integers.  $O(\log N)$ rounds.


%%%%%

\section{Monitors}

\input{buffer} % \framebox{Implement buffer (extending slot).}

\input{consumerM} % \framebox{Consumer (Bear \& honeybees) (A 4.36)}

\input{menWomenM}

\input{mod3M}

\framebox{Cigarette smokers problem, A4.27.}

\framebox{Sleeping barber (A Sec 5.2.5)} ??

\framebox{Savings account A 5.8 / B Ex 1.1} easy

\framebox{Resource allocation (B Ex 1.?)} ??

\framebox{Producer (Dining savages (A 5.12))}


 
\framebox{Deli} hard

\framebox{Sleeping don} hard

\framebox{Unisex bathroom} easy

\framebox{***} roller coaster ?

\framebox{***} SJF

\framebox{***} One lane bridge??


%%%%%

\section{Patterns}

%\section{Recursive parallelism}

\input{treeSum} % Sum of elements in a tree. -- one or two processes forked.

\framebox{evaluation of arithmetic expressions}

\framebox{MSort}

\input{fibonacci} % Sheet 4

\framebox{Matrix multiplication by blocks}

\framebox{Readers \& writers}

%%%%%

%\section{Bag of tasks}

% \framebox{Eight queens} -- MSc assignment?

\input{adaptive} % Sheet 4 \framebox{Adaptive quadrature}

\framebox{TSP}

\framebox{Suppose the controller is a bottleneck.}  What can you do?

\input{distBagTasks}
 % \framebox{Distributed bag of tasks:} each process has its own bag, as a
 % subprocess, which may pass tasks either to its own process or to another
 % process.  Leads to distributed termination. ***

\framebox{A 3.28 (generating primes)}

\framebox{***} map-reduce \url{http://labs.google.com/papers/mapreduce.html}?.  Optional:
some features from Secs 3, 4.  Polymorphic.

\framebox{Countdown?}

%%%%%

\section{Semaphores}

\input{consumerS} % sheet 4

\input{menWomenS} % sheet 4

\input{readersWritersBaton} % sheet 4

\input{boundedBuffSemaphore} % Bounded buffer using semaphores, sheet 4.

\framebox{Adapt barrier synchronisation} so that each node supplies some data,
and they get back some function (fold) of that data.

\framebox{(Most examples from monitors section).}

\framebox{Combining barrier, A Sec 3.4.2?}

\framebox{Tournament barrier A 3.14 ?}

\framebox{A 4.4 (precedence)}


\framebox{Bounded counting semaphore (using two semaphores)}

\framebox{Implement a counting semaphore using a binary semaphore.}

\framebox{Atomic broadcast (A 4.16)}

\framebox{Incorrect solution to readers \& writers problem} -- explain problem

\framebox{The water molecule problem (A 4.25)}

\framebox{Unisex bathroom (A 4.30).}

\framebox{Search / insert / delete A 4.32}

\framebox{(4.35, 4.36, 4.37)}

\input{mod3S} % - done in lectures

\framebox{Cigarette smokers - A4.27}

\framebox{One lane bridge - A4.31}



%%%%%

\section{Mixed synchronisation}

\input{account}

\input{twoTypesProc} (= Unisex bathroom)

\end{document}
