\begin{question}
\begin{enumerate}
\item
Write a definition for a process
\begin{scala}
  def sumStreams(in1: ?[Int], in2: ?[Int], out: ![Int]) = proc{ ... }
\end{scala}
that repeatedly inputs a value from each of its input ports (in either order),
and outputs the sum on the output port.  (Functional programmers might like to
think of this process as implementing \SCALA{zipwith (+)}.)

\item
The Fibonacci numbers \SCALA{fibs} (0, 1, 1, 2, 3, 5, \ldots) can be defined
using the equation
\begin{scala}
  fibs = 0 : sumStreams( fibs, 1:fibs )
\end{scala}
(i.e.\ the stream that starts with 0, and continues with the result of summing
the stream \SCALA{fibs} and the result of pre-pending \SCALA{fibs} with~1). 

Design and implement a circuit with signature
\begin{scala}
  def Fibber(out: ![Int]) = ...
\end{scala}
that outputs the Fibonacci numbers on \SCALA{out}, making use of the above
equation.  Draw a picture to explain your design.  Produce a simple test rig
to test your implementation.
\end{enumerate}
\end{question}

%%%%%

\begin{answer}
We'll produce a system as illustrated below.
\begin{verbatim}
                      fwd
    |---->prefix 0-----------|  
    |                        |
    |                        |
    |                back1   V    out
sum |      prefix 1<-------tee3---->
    |         |              |
    |    back2|              |back3
    |         |              |
    |         V              |
    ----sumStreams<-----------                           
\end{verbatim}
\begin{scala}
import ox.CSO._

object Fibs{

  // Three-way T component
  def tee3[T](in: ?[T], out1: ![T], out2: ![T], out3: ![T]) = proc{
    repeat{
      val v = in?; 
      (proc{ out1!v } || proc{ out2!v } || proc{ out3!v })();
    }
    in.close; out1.close; out2.close; out3.close;
  } 

  // Input values from each input channel, and output their sum
  def sumStreams(in1: ?[Int], in2: ?[Int], out: ![Int]) = proc{
    var x=0; var y=0;
    repeat{
      (proc{ x = in1? } || proc{ y = in2? })();
      out!(x+y); 
    }
    in1.close; in2.close; out.close;
  }

  // From lectures
  def prefix[T](first:T, in: ?[T], out: ![T]) = proc{
    out!first;
    repeat{ out!(in?); };
    in.close; out.close;
  }

  val sum, fwd, back1, back2, back3 = OneOne[Int];

  def Fibber(out: ![Int]) = (
    prefix(0, sum, fwd) || tee3(fwd, out, back1, back3) ||
    prefix(1, back1, back2) || sumStreams(back2, back3, sum)
  );

  // Test rig
  def TestRig = {
    val out = OneOne[Int];
    Fibber(out) || ox.cso.Components.console(out);
  }

  def main(args : Array[String]) = TestRig()
}
\end{scala}
\end{answer}
