\begin{nontutequestion}
\Programming
\begin{enumerate}
\item
Write a definition for a component
%
\begin{scala}
def zipwith[L,R,O](f:(L, R)=>O)(lin:?[L], rin:?[R], out:![O]) = proc{ ... }
\end{scala}
%
which repeatedly inputs values \SCALA{l} and \SCALA{r} on \SCALA{lin} and
\SCALA{rin}, respectively, and outputs \SCALA{f(l,r)} on \SCALA{out}. 

\item
An \emph{integrator} is a process with signature
%
\begin{scala}
def integrator(in: ?[Int], out: ![Int]) = ...
\end{scala}
that repeatedly inputs on~\SCALA{in}, and outputs on~\SCALA{out} the sums of
the inputs so far.  That is, if it receives the inputs \SCALA{x1, x2, x3,
  ...}, it outputs \SCALA{x1, x1+x2, x1+x2+x3, ...}.  Design and implement an
integrator as a circuit using \SCALA{zipwith} and other common components.
Draw a diagram to explain your design.  Produce a simple test rig to test the
integrator. 
\end{enumerate}
\end{nontutequestion}

%%%%%

\begin{nontuteanswer} 
\begin{scala}
def zipwith[L,R,O](f:(L, R)=>O)(lin:?[L], rin:?[R], out:![O]) = proc{
  var l = null.asInstanceOf[L]; 
  var r = null.asInstanceOf[R];
  repeat { 
    (proc { l = lin? } || proc { r = rin? })(); 
    out!f(l, r) 
  };   
  lin.close; rin.closein; out.closeout
}
\end{scala}
%
In fact, this is defined as \SCALA{ox.cso.Components.zipwith}.

For the integrator, we'll use a circuit like before (where the first component
is a \SCALA{zipwith} using adition.
%
\begin{verbatim}
                in                       out
 [x,y,z,...] >------>|\    mid     /|-------> [x,x+y,x+y+z,...]
                     |+}--------->{ |
                 +-->|/            \|--+
             addl|                     |back
                 +------<prefix 0<-----+
\end{verbatim}
%
\begin{scala}
import ox.CSO._, ox.cso.Components

object Integrator{

  def integrator(in: ?[Int], out: ![Int]) = {
    val mid, back, addl = OneOne[Int]
    (  Components.zipwith ((x:Int, y:Int)=>x+y) (in, addl, mid)
    || Components.tee (mid, List(out, back))
    || Components.prefix(0)(back, addl)
    )
  }

  // Produce stream of nats
  def nats(out: ![Int]) = proc("nats"){
    var n = 0;
    repeat{ out!n; n+=1; sleep(200) }
  }

  // Test rig, using the above
  def TestRig = {
    val in, out = OneOne[Int];
    ( integrator(in, out) || nats(in) || Components.console(out) )
  }

  def main(args : Array[String]) = TestRig()
}
\end{scala}
The test rig provides the integrator with naturals, and we can check that we
get the triangular numbers out.
\end{nontuteanswer}
