\begin{question}
\Programming\ 
Recall the men-women problem from question~\ref{Q:menWomenM}.  
% Recall the following synchronisation problem from the previous
% sheet.  There are two types of client process, which we shall call \emph{men}
% and \emph{women}.  These processes need to pair off for some purpose, with
% each pair containing one process of each type.  Design a class to support
% this, with public methods
% \begin{scala}
%   def manSync(me: String): String = ...
%   def womanSync(me: String): String = ...
% \end{scala}
% %
% Each process should pass its name to the monitor, and receive back the name of
% its partner.  
Implement a class to solve this problem, using semaphores internally.
Test your code.
\end{question}

%%%%%

\begin{answer}
My code is below.  Each round proceeds as follows.
%
\begin{enumerate}
\item[1.] A man waits (on |manWait|) for the previous round to complete, writes
his name, signals to a woman (on |womanWait|), and waits (on |manWait2|) for a
signal back;

\item[2.] A woman waits (on |womanWait|) for a man, reads his name, stores her
name, and signals back (on |manWait2|);

\item[3.] The man reads her name, and signals (on |manWait|) to a man waiting for
the next round.
\end{enumerate}
%
\begin{scala}
/** A solution using semaphores. */
class MenWomenSemaphore extends MenWomen{
  /** Name of the current (or last) woman. */
  private var woman = "Alice"

  /** Name of the current (or last) man. */
  private var man = "Bob"

  /** Semaphore where a man waits initially. */
  private val manWait = MutexSemaphore()

  /** Semaphore where a woman waits until signalled. */
  private val womanWait = SignallingSemaphore()

  /** Semaphore where the current man waits for a woman. */
  private val manWait2 = SignallingSemaphore()

  // The above three semaphores collectively ensure mutual exclusion.

  def manSync(me: String): String = {
    manWait.down    // wait my turn (1)
    man = me         // store my name
    womanWait.up    // signal to a woman at (2)
    manWait2.down   // wait for an acknowledgement (3)
    val her = woman // get her name
    manWait.up       // signal to the next man at (1)
    her
  }

  def womanSync(me: String): String = {
    womanWait.down                // wait for a man (2)
    woman = me; val him = man // store my name, get his
    manWait2.up                    // signal to him at (3)
    him
  }
}
\end{scala}

There are lots of other solutions.

Testing can be done identically to as on previous sheets.
\end{answer}
