\begin{question}
\Programming\ (Based on Andrews, Exercise 7.11.)
%
Cars coming from the north and the south arrive at a one-lane bridge.  Cars
heading in the same direction can cross the bridge at the same time, but cars
heading in the opposite direction cannot.

Develop a server process to manage use of the bridge.  Assume that cars are
client processes.  Describe the protocol between the cars and the server.
\end{question}

%%%%%

\begin{answer}
The protocol is that cars send a Arrive message when they arrive, wait for
an acknowledgement (a Unit), cross the bridge, then send a Leave message.
The server allows cars to proceed if there is no car crossing or waiting to
cross in the opposite direction; otherwise the car queues.
%
\begin{scala}
import ox.CSO._;

object OneLaneBridge{
  type ClientId = Int;
 
  abstract class Request;
  case class Arrive(fromN:Boolean, c:ClientId) extends Request;
  case class Leave(c:ClientId) extends Request;

  // The server
  def Server(req: ?[Request], resp: Seq[![Unit]]) = proc{
    var currentN = true; // is current priority from north?    
    // Queue of cars queuing in the opposite direction  
    var queue = new scala.collection.mutable.Queue[ClientId] 
    // Queues going in the current direction, now queueing
    var nextQueue = new scala.collection.mutable.Queue[ClientId] 
    var count = 0; // number of cars currently crossing

    // Convert direction to string
    def showDir(isN:Boolean) = if(isN) "North" else "South";

    repeat{
      req? match{
	case Arrive(fromN, c) => {
	  if(fromN==currentN){ // Car going in current direction
	    if(queue.isEmpty){ // Allow to proceed
	      resp(c)!(); count += 1;
	      println("Car "+c+" arrives from "+
		      showDir(fromN)+" and proceeds")
	    } else { // Must wait
	      nextQueue += c;
	      println("Car "+c+" arrives from "+
		      showDir(fromN)+" and queues")
	    }
	  }
	  else if(count==0){ 
            // Reverse direction, allow to proceed
	    currentN = fromN; resp(c)!(); count = 1;
	    println("Car "+c+" arrives from "+
		    showDir(fromN)+" and proceeds")
	  } else { // Add to queue
	    queue += c;
	    println("Car "+c+" arrives from "+
		    showDir(fromN)+" and queues")
	  }
	}
	case Leave(c) => { // Detect a car has left
	  count -= 1;
	  println("Car "+c+" leaves");
	  if(count==0 && !queue.isEmpty){ // Reverse direction
	    currentN = !currentN;
	    // Set the queue going
	    while(!queue.isEmpty){ 
	      val c = queue.dequeue; resp(c)!(); count += 1;
	      println("Car "+c+" dequeued, and continues "+
		      showDir(currentN))
	    }
	    // Swap the queues
	    val temp = queue; 
            queue = nextQueue; nextQueue = queue;
	  }
	}
      }
    }
  }

  val random = new scala.util.Random;

  // A car
  def Car(me: Int, req: ![Request], resp: ?[Unit]) = proc{
    repeat{
      sleep(random.nextInt(8000));
      // Arrive from random direction
      req!Arrive(random.nextInt(2)==0, me); 
      resp?; // Wait for permission 
      sleep(random.nextInt(2000)); // Cross the bridge
      req!Leave(me); // And away
    }
  }

  // Put system together
  val numCars = 5;
  val req = ManyOne[Request];
  val resp = OneOne[Unit](numCars);

  def Cars = 
    || ( for(i <- 0 until numCars) yield Car(i, req, resp(i)) )

  def System = Server(req, resp) || Cars

  def main(args:Array[String]) = System();
}
\end{scala}
\end{answer}
