\begin{question}
A particular system contains two types of client process, which we shall call
A and~B.  They share a resource.  However, the resource may be used by only
one type of client process at a time: if a client tries to access the
resource while a client of the other type is using it, then it must block
until all clients of the other type finish.  There is no limit on the number
of clients of a single type that may use the resource at one time.

The aim of this question is to enforce the above requirement using the three
different types of concurrency primitive: message passing, monitors and
semaphores.  In each part of the question you should use only the concurrency
primitive mentioned.

You will probably find that your code for types A and B are very similar.  If
so, you should give the code for one type fully, but you may elide the code
for the other type.

%%%%%

\begin{enumerate}
\item
Implement a server process to enforce the above requirement.  Clients of type
A and B should synchronise with the server on the ManyOne channels
\SCALA{AEnter} and \SCALA{BEnter}, respectively, before using the resource;
these communications should be blocked when clients of the other type are
using the resource.  The clients should synchronise with the server on the
ManyOne channels \SCALA{ALeave} and \SCALA{BLeave}, respectively, when they
finish using the resource.  Hence the signature of the server is:
%
\begin{scala}
def Server(AEnter: ?[Unit], BEnter: ?[Unit], 
           ALeave: ?[Unit], BLeave: ?[Unit]) = proc{ ... }
\end{scala}
\marks{6}

%%%%%

\item
Now implement a monitor with the following signature to enforce the
requirement:
%
\begin{scala}
object Monitor{
  def AEnter = ...
  def BEnter = ...
  def ALeave = ...
  def BLeave = ...
}
\end{scala}
\marks{7}

\item
Finally, use semaphores to implement an object with the same signature as in
the previous part to enforce the requirement.  \marks{12}
\end{enumerate}
\end{question}

%%%%%%%%%%

\begin{answer}
{\bf Part a}
\begin{scala}
def Server(AEnter: ?[Unit], BEnter: ?[Unit], 
           ALeave: ?[Unit], BLeave: ?[Unit])
= proc{
  var typeA = 0; var typeB = 0; 
  // Number of type A, type B currently in

  serve(
    (typeB==0 &&& AEnter) --> { AEnter?; typeA += 1; }
    | (typeA==0 &&& BEnter) --> {BEnter?; typeB += 1; }
    | ALeave --> { ALeave?; typeA -= 1; }
    | BLeave --> { BLeave?; typeB -= 1; }
  )
}
\end{scala}
[Quite easy.]

%%%%%

{\bf Part b}
%
\begin{scala}  
object Monitor{
  private var typeA = 0; private var typeB = 0; 
  // Number of type A, type B currently in

  def AEnter = synchronized{
    while(typeB>0) wait();
    typeA += 1;
  }

  def BEnter = synchronized{
    while(typeA>0) wait();
    typeB += 1;
  }

  def ALeave = synchronized{
    typeA -= 1;
    if(typeA==0) notifyAll(); // wake the others up
  }

  def BLeave = synchronized{
    typeB -= 1;
    if(typeB==0) notifyAll(); // wake the others up
  }
}
\end{scala}
[Reasonably straightforward.]

%%%%%

{\bf Part c}
%
\begin{scala}
object Monitor{
  private var typeA = 0; private var typeB = 0; 
  // Number of type A, type B currently in

  private var AWaiting = 0; private var BWaiting = 0;
  // Number of A, B currently waiting

  private val AEntry, BEntry = new Semaphore;
  AEntry.down; BEntry.down
  // Semaphores for A, B to wait on

  private val mutex = new Semaphore;

  // Unblock somebody, if anybody waiting
  private def Signal = {
    if(typeA==0 && BWaiting>0){ BEntry.up; }
    else if (typeB==0 && AWaiting>0){ AEntry.up; }
    else mutex.up;
  }

  def AEnter = {
    // Await typeB==0
    mutex.down;
    if(typeB>0){  // wait to be allowed to enter
      AWaiting += 1; mutex.up; AEntry.down; AWaiting -= 1; 
    }      
    typeA += 1;
    Signal;
  }

  def BEnter = {
    // Await typeA==0
    mutex.down;
    if(typeA>0){  // wait to be allowed to enter
      BWaiting += 1; mutex.up; BEntry.down; BWaiting -= 1; 
    }      
    typeB += 1;
    Signal; 
  }

  def ALeave = {
    mutex.down;
    typeA -= 1;
    Signal;
  }

  def BLeave = {
    mutex.down;
    typeB -= 1;
    Signal; 
  }
}
\end{scala}
[Quite difficult.]
\end{answer}
